---
title: "BiocAnnotationLab: developing new approaches to genomic annotation curation"
shorttitle: "Renewing annotation strategies in Bioconductor"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{BiocAnnotationLab: developing new approaches to genomic annotation curation}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Motivation

Bioconductor's annotation resources are extensive and can be challenging to maintain
for various reasons.  This package collects information relevant to 

- reducing complexity of annotation management
- taking advantage of new approaches to data representation
- supporting "self-service" solutions for those needing rapid revision of annotation resources

Basic facts:

- `GO.db`
    - Package is downloaded by over 20000 distinct IPs per month
    - It is powered by a 70MB SQLite database with 14 tables following a bespoke schema
    - The production of the GO.db package is mixed with production of numerous other annotation
packages in the BioconductorAnnotationPipeline system.
    - A parquet-based representation with equivalent content can fit in 7MB of disk.  See
[GO.db3](https://github.com/vjcitn/GO.db3) and note the OBO-parquet transformation
resource at [BiocGOprep](https://github.com/vjcitn/BiocGOprep).
    - Yet another approach to Gene Ontology (and indeed all ontologies at OBOFoundry)
leverages semantic SQL -- see [ontoProc2](https://vjcitn.github.io/ontoProc2/), in development.

- `org.*.*.db`
    - All the org.*.eg.db are based on curation of resources from NCBI
    - These do not need to be carved up for different organisms; see the
[RNCBIgene](https://vjcitn.github.io/RNCBIGene/) package in development.
    - Project-specific resources like SGD, TAIR, PLASMO had special casing
in the pipeline, and this needs a fresh, maintainable approach

- `TxDb.*.*...`
    - The architect is still with the project
    - Tooling to move from SQLite to parquet is available and produces 7x reduction
in footprint
    - Upstream tasks are well-documented, but should be reviewed regularly

# Gene Ontology: approaches to curation

## GO.db

Based on download counts, GO.db is a very popular resource in Bioconductor.
Here's a basic use case, taken from the sources of goana.R in limma.

```
TERM <- suppressMessages(AnnotationDbi::select(GO.db::GO.db,keys=GOID,columns="TERM"))
```

Try it out:
```{r setup}
GOID = c("GO:0001960", "GO:0001961", "GO:0010803", "GO:0060334", "GO:0060338", 
"GO:0070099", "GO:0070103", "GO:0070107", "GO:0070758", "GO:1900234", 
"GO:1902205", "GO:1902211", "GO:1902214", "GO:1902226", "GO:1903881", 
"GO:2000446", "GO:2000492", "GO:2000659")
library(GO.db)
TERM <- suppressMessages(AnnotationDbi::select(GO.db::GO.db,keys=GOID,columns="TERM"))
head(TERM)
```

GO.db is self-descriptive.
```{r lkself}
GO.db
```

We can learn about the underlying database.

```{r lkconn}
file.size(slot(GO_dbconn(), "dbname"))
DBI::dbListTables(GO_dbconn())
```

GO.db presents environments with traversals of the ontological
hierarchy.

```{r lkhi1}
get("GO:0001959", GO.db::GOBPCHILDREN)
```

## GO.db3: a parquet-based approach

In pursuit of a proof of concept, packages GO.db2 (SQLite-based) and GO.db3
have been produced.  GO.db2 uses a simplified schema for the tables; GO.db3
uses parquet representations of GO.db's tables.

To avoid conflict with dplyr's `select`, GO.db3 defines `select3` to
emulate AnnotationDbi::select.

```{r chk3}
library(GO.db3)
chk = select3("GO.db3",keys=GOID, columns=c("GOID", "TERM"), keytype="GOID")
head(chk)
```

We can learn about the underlying resources, in `extdata/go323` and
observe a sharp reduction in size of data footprint.

```{r lkszzz}
pfns = dir(system.file("extdata", "go323", package="GO.db3"), full=TRUE)
head(basename(pfns))
sum(file.size(pfns))
```

We wrap the hierarchy-oriented environments in functions for now.
The environments are computed "on the fly"; only a few have been produced.

```{r lkhi2}
get("GO:0001960", GO.db3::GOBPPARENTS())
```

## ontoProc2 for GO: using Semantic SQL

The [INCAtools semantic SQL](https://github.com/INCATools/semantic-sql) project
converts ontologies in OWL or OBO formats to SQLite.  The README remarks that

```
SQLite provides many advantages

 - files can be downloaded and subsequently queried without network latency
 - compared to querying a static rdf, owl, or obo file, there is no startup/parse delay
 - robust and performant
 - excellent support in many languages
Although the focus is on SQLite, this library can also be used for other 
DBMSs like PostgreSQL, MySQL, Oracle, etc
```

Here's how we can use a Semantic SQL representation of GO with ontoProc2.
The `GOID` vector was defined above.

```{r useo2}
library(ontoProc2)
gosem = retrieve_semsql_conn("go")
chk1 = dplyr::tbl(gosem, "statements") |> 
  dplyr::filter(subject %in% GOID, predicate=="rdfs:label") |> 
  dplyr::select(subject, value)
head(chk1)
```

Note that the full representation of all facts and relationships in GO
entails a large footprint.

```{r lkfp}
file.size(slot(gosem, "dbname"))
```

## Upshots

- AnnotationDbi's production of GO.db involves a complex pipeline with
processes intermingled to address multiple objectives.
- AnnotationDbi's "select" method is in conceptual conflict with
the widely used select method of dplyr.
- [GO.db3](https://github.com/vjcitn/GO.db3?tab=readme-ov-file#godb3-is-an-experimental-package-that-partially-emulates-godb) is a near-complete functionally compatible replacement for GO.db.
- Production of GO.db3 is independent of all other annotation production
processes; parquet production from OBO is documented in [BiocGOprep](https://github.com/vjcitn/BiocGOprep).
- `GO.db3::select3` is provided to emulate `AnnotationDbi::select`, but other
interrogation interfaces based on the parquet representation should
be devised.  BiocGOprep produces parquet files following GO.db's schema; this
is not essential but was adopted to simplify producing compatibility with legacy approaches.
- Both GO.db and GO.db3 present only a very limited view of the relationships
encoded in Gene Ontology; a complete view is made available for SQL-based
interrogation via ontoProc2.
- ontoProc2 helps provide full programmatic access to any ontology
managed in the Semantic SQL project.

## Next steps

- Community members are invited to comment on implications of
strategic choices on annotation curation for their work.
- GO.db3 should be brought to verifiable functional equivalence with GO.db.
- Other annotation elements should be looked at for opporunities to
use more efficient representations and interfaces.
